Estimating Predictive Rate-Distortion via Neural Variational Inference

### 2019-03-06 paper.tex
code

The \emph{distortion} is the loss in predictive accuracy when predicting from $Z$, relative to optimal prediction from the full past $\overleftarrow{X}$.
In the Information Bottleneck formalization, this
%the distortion measure is computed based on the entropy:
%\begin{equation}
%H[\overrightarrow{X}|Z] - H[\overrightarrow{X}|\overleftarrow{X}]
%\end{equation}
%This
is equivalent to the amount of mutual information between past and future that is not captured by $Z$:
\begin{equation}
	I[\overleftarrow{X}, \overrightarrow{X}|Z]
\end{equation}
The distortion measure satisfies the relation
\begin{equation}
	I[\overleftarrow{X}, \overrightarrow{X}|Z] = I[\overleftarrow{X}, \overrightarrow{X}] - I[\overrightarrow{X}, Z]
\end{equation}
so that, for a fixed process $(X_t)_t$ minimizing the distortion is equivalent to maximizing $I[\overrightarrow{X}, Z]$.

The rate-distortion tradeoff then chooses $Z$ to minimize rate at bounded distortion:
$$\min_{Z : I[\overleftarrow{X}, \overrightarrow{X}|Z] \leq d} I[Z, \overleftarrow{X}]$$
, or -- equivalently -- minimize distortion at bounded rate.
$$\min_{Z :  I[Z, \overleftarrow{X}] \leq d}  I[\overleftarrow{X}, \overrightarrow{X}|Z] $$
Equivalently, for each  $\beta \geq 0$, we study the problem 
\begin{equation}\label{eq:ib}
	\max_{Z} \left( I[\overrightarrow{X}, Z] - \beta \cdot I[Z, \overleftarrow{X}] \right)
\end{equation}
where the scope of the maximization is the class of all random variables $Z$ such that $Z - \overleftarrow{X} - \overrightarrow{X}$ is a Markov chain. %, that is, $Z$ is independent of the future given the past.
It is not hard to see that this optimized by a trivial $Z$ with zero rate and maximal distortion when $\beta \geq 1$, so we will be concerned with the situation where $\beta \in [0,1]$.

@@

Let us assume that an \emph{encoder} $\phi$ is given, expressed in a family of function approximators. 
The encoder stochastically transforms observation sequences $x_{-M\dots -1}$ into a random code $Z \in \mathbb{R}^N$: 
\begin{equation}
Z = \phi(x_{-M\dots -1})
\end{equation}
distributed according to $Z \sim P_\phi(Z|x_{-M\dots -1})$.
From this, one can obtain the conditional distribution over future observations via Bayes' rule: %, using the Markov condition: \rljf{Show some intermediate steps here}
\begin{equation}
	\label{eq:cond-fut}
	\begin{split}
	P(X_{0\dots M-1}|Z) &= \frac{p(X_{0\dots M-1}, Z)}{p(Z)} \\
	&= \frac{\E_{X_{-M \dots -1}} P_\phi(X_{0\dots M-1}, Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})} \\
	&=^{(*)} \frac{\E_{X_{-M \dots -1}}   P(X_{0\dots M-1}|X_{-M\dots-1})     P_\phi(Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})} \\
	&= \frac{\E_{X_{-M \dots M-1}} P_\phi(Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})}
	\end{split}
\end{equation}
where $(*)$ uses the Markov condition.
However, neither of the two expectations is tractable,  as they require summation over exponentially many sequences.
If $\phi$ is given by a powerful function approximating family such as recurrent neural networks, algorithms (e.g., dynamic programming) to compute this sum efficiently are not available.

@@

That is, we introduce $\psi$ and $q$ as \emph{variational approximations} \citep{blei-variational-2016} to the true Bayes optimal decoder and marginal. \rljf{Expand last sentence into a paragraph}

@@

\paragraph{OCE}
As discussed above, OCE overestimates structure in the process due to overfitting.
To address this, and to enable fair comparison between OCE and our neural method, we follow the evaluation method used for NPRD, evaluating rate and distortion on held-out data:
\rljf{Describe these in prose too and what they mean}
We estimate rate using the variational bound that we stated in Proposition~\ref{ineq2}:
\begin{equation}
\frac{1}{N} \sum_{i=1}^N \operatorname{D_{KL}}(\infdiv{q(\cdot|x_i)}{s(\cdot)})
\end{equation}
We estimate the prediction loss on the future observations as the empirical cross-entropy, i.e., the variational bound stated in Proposition~\ref{ineq1}:
\begin{equation}
\frac{1}{N} \sum_{i=1}^N \E_{z \sim q(\cdot|x_i)} \log r(y_i|z).
\end{equation}
Thanks to (\ref{ineq1}-\ref{ineq2}), these quantities provide upper bounds, up to sampling error introduced by finiteness of the held-out data.
Again, this kind of error does not bias the results in either direction, unlike overfitting, which introduces a systematic bias.

@@

\begin{table}
\begin{tabular}{l||lll}
	            & Section~\ref{sec:tractable} \\ \hline\hline
	Embedding Dimension & 50, 100 \\
	LSTM Dimension & 32 \\ \hline
%	LSTM Layers & 1 \\ \hline
	Dropout rate & 0.0, 0.1, 0.2 \\
	Input Dropout & 0 \\ \hline
	Adam Learning Rate & 0.0001, 0.0005, 0.001 (4 times), 0.002, 0.004 \\
	Batch Size & 16, 32, 64 \\ \hline
	Flow Length & 1, 2 \\
	Flow Type & DSF, DDSF \\
	Flow Dimension & 32, 64, 128, 512 \\
	Flow Layers & 2 \\
\end{tabular}
	\caption{NPRD Hyperparameters}\label{tab:nprd-hyperparameters}
\end{table}

@@

%\includegraphics[width=0.45\textwidth]{code/figures/even-beta-mem.pdf}
	\caption{Rate-Distortion for the Even Process and the Random Insertion Process. Gray lines: Analytical curves. Red dots: Multiple runs of NPRD. Red line: Tradeoff curve computed from NPRD runs. Blue: OCE for $M\leq 5$. }\label{fig:even}
\end{figure*}

@@

\begin{figure*}
%\includegraphics[width=0.45\textwidth]{code/figures/repeat2-ee-mem.pdf}
\includegraphics[width=0.45\textwidth]{code/figures/repeat3-ee-mem-frontier.pdf}
	\caption{Rate-Distortion for the \textsc{Copy3} process. We show NPRD samples, and the resulting upper bound in red. The gray line represents the anaytical curve.}\label{fig:repeat}
\end{figure*}
%  \rljf{Red not green? Also y axis should be Memory, not Memories. Also let's label as Memory (nats) throughout (or bits), and Excess Entropy on the x axis throughout.}

@@

There is the related question of how much memory is needed to predict.

@@

>>>>

The \emph{distortion} is the loss in predictive accuracy when predicting from $Z$, relative to optimal prediction from the full past $\overleftarrow{X}$.
In the Information Bottleneck formalization, this
%the distortion measure is computed based on the entropy:
%\begin{equation}
%H[\overrightarrow{X}|Z] - H[\overrightarrow{X}|\overleftarrow{X}]
%\end{equation}
%This
is equivalent to the amount of mutual information between past and future that is not captured by $Z$:
\begin{equation}
	I[\overleftarrow{X}, \overrightarrow{X}|Z]
\end{equation}
Due to the Markov condition, the distortion measure satisfies the relation
\begin{equation}
	I[\overleftarrow{X}, \overrightarrow{X}|Z] = I[\overleftarrow{X}, \overrightarrow{X}] - I[\overrightarrow{X}, Z]
\end{equation}
For a fixed process $(X_t)_t$, minimizing the distortion is equivalent to maximizing the mutual information between the code and the future $I[\overrightarrow{X}, Z]$. We will refer to $I[\overrightarrow{X}, Z]$ as the \emph{predictiveness} of the code $Z$.

The rate-distortion tradeoff then chooses $Z$ to %minimize rate at bounded distortion:
%$$\min_{Z : I[\overleftarrow{X}, \overrightarrow{X}|Z] \leq d} I[Z, \overleftarrow{X}]$$
%, or -- equivalently --
 minimize distortion at bounded rate:
\begin{equation}
\min_{Z :  I[Z, \overleftarrow{X}] \leq d}  I[\overleftarrow{X}, \overrightarrow{X}|Z] 
\end{equation}
or -- equivalently -- maximize predictiveness at bounded rate:
\begin{equation}
\max_{Z :  I[Z, \overleftarrow{X}] \leq d}  I[Z, \overrightarrow{X}] 
\end{equation}
Equivalently, for each  $\beta \geq 0$, we study the problem 
\begin{equation}\label{eq:ib}
	\max_{Z} \left( I[\overrightarrow{X}, Z] - \beta \cdot I[Z, \overleftarrow{X}] \right)
\end{equation}
where the scope of the maximization is the class of all random variables $Z$ such that $Z \rightarrow \overleftarrow{X} \rightarrow \overrightarrow{X}$ is a Markov chain. %, that is, $Z$ is independent of the future given the past.

The objective (\ref{eq:ib}) is equivalent to the Information Bottleneck \citep{tishby-information-1999}, applied to the past and future of a stochastic process.

When $\beta \geq 1$, then (\ref{eq:ib}) is always optimized by a trivial $Z$ with zero rate and zero predictiveness, which follows from the Data Processing Inequality.
When $\beta = 0$, any lossless code optimizes the problem.
Therefore, we will be concerned with the situation where $\beta \in (0,1)$.

@@

Let us assume that an \emph{encoder} $\phi$ is given, expressed in a family of function approximators. 
The encoder stochastically transforms observation sequences $x_{-M\dots -1}$ into a random code $Z \in \mathbb{R}^N$: 
\begin{equation}
Z = \phi(x_{-M\dots -1})
\end{equation}
distributed according to $Z \sim P_\phi(Z|x_{-M\dots -1})$.
From this, one can obtain the conditional distribution over future observations via Bayes' rule: %, using the Markov condition: 
\begin{equation}
	\label{eq:cond-fut}
	\begin{split}
	P(X_{0\dots M-1}|Z) &= \frac{p(X_{0\dots M-1}, Z)}{p(Z)} \\
	&= \frac{\E_{X_{-M \dots -1}} P_\phi(X_{0\dots M-1}, Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})} \\
	&=^{(*)} \frac{\E_{X_{-M \dots -1}}   P(X_{0\dots M-1}|X_{-M\dots-1})     P_\phi(Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})} \\
	&= \frac{\E_{X_{-M \dots M-1}} P_\phi(Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})}
	\end{split}
\end{equation}
where $(*)$ uses the Markov condition.
However, neither of the two expectations is tractable,  as they require summation over exponentially many sequences.
If $\phi$ is given by a powerful function approximating family such as recurrent neural networks, algorithms (e.g., dynamic programming) to compute this sum efficiently are not available.

@@

That is, we introduce $\psi$ and $q$ as \emph{variational approximations} \citep{blei-variational-2016} to the true Bayes optimal decoder and marginal.

@@

\paragraph{OCE}
As discussed above, OCE overestimates structure in the process due to overfitting.
To address this, and to enable fair comparison between OCE and our neural method, we follow the evaluation method used for NPRD, evaluating rate and distortion on held-out data:
We estimate rate using the variational bound that we stated in Proposition~\ref{ineq2}:
\begin{equation}
\frac{1}{N} \sum_{i=1}^N \operatorname{D_{KL}}(\infdiv{q(\cdot|x_i)}{s(\cdot)})
\end{equation}
We estimate the prediction loss on the future observations as the empirical cross-entropy, i.e., the variational bound stated in Proposition~\ref{ineq1}:
\begin{equation}
\frac{1}{N} \sum_{i=1}^N \E_{z \sim q(\cdot|x_i)} \log r(y_i|z).
\end{equation}
Thanks to (\ref{ineq1}-\ref{ineq2}), these quantities provide upper bounds, up to sampling error introduced by finiteness of the held-out data.
Again, this kind of error does not bias the results in either direction, unlike overfitting, which introduces a systematic bias.

@@

\begin{table}
\begin{tabular}{l||lll}
	& Section~\ref{sec:tractable} &   POS & Words\\ \hline\hline
	Embedding Dimension & 50, 100 &       &  150      \\
	LSTM Dimension & 32           &       &    512     \\ \hline
%	LSTM Layers & 1 \\ \hline
	Dropout rate & 0.0, 0.1, 0.2  &       &      0.4     \\
	Input Dropout & 0  &       &        0.2             \\ \hline
	Adam Learning Rate & 0.0001, 0.0005, 0.001 (4 times), 0.002, 0.004  &       &       0.00005, 0.0001, 0.0005, 0.001     \\
	Batch Size & 16, 32, 64  &       &       32          \\ \hline
	Flow Length & 1, 2  &       &          4              \\
	Flow Type & DSF, DDSF  &       &       DSF, DDSF             \\
	Flow Dimension & 32, 64, 128, 512  &       &    ???        \\
	Flow Layers & 2  &       &                 ?????          \\
\end{tabular}
	\caption{NPRD Hyperparameters}\label{tab:nprd-hyperparameters}
\end{table}

@@

%\includegraphics[width=0.45\textwidth]{code/figures/even-beta-mem.pdf}
	\caption{Rate-Distortion for the Even Process (left) and the Random Insertion Process (right). Gray lines: Analytical curves. Red dots: Multiple runs of NPRD. Red line: Tradeoff curve computed from NPRD runs. Blue: OCE for $M\leq 5$. }\label{fig:even}
\end{figure*}

@@

\begin{figure*}
%\includegraphics[width=0.45\textwidth]{code/figures/repeat2-ee-mem.pdf}
\includegraphics[width=0.45\textwidth]{code/figures/repeat3-info.pdf}
	\caption{Rate-Distortion for the \textsc{Copy3} process. We show NPRD samples, and the resulting upper bound in red. The gray line represents the anaytical curve.}\label{fig:repeat}
\end{figure*}

@@

\cite{smith-effect-2013}

\cite{Bar:2007,Clark:2016}
There is the related question of how much memory is needed to predict.

@@

\cite{futrell-noisy-context-2017}
